Global Distribution of Unintended Pregnancies and Contraceptive Usage

### 2024-03-14 code/analysis-code/statistical-analysis.qmd
Statistical Analysis Update
I cleaned this section to only include the one outcome we chose with models for each of the 7 predictors. I also added RMSE values for 2 models to begin comparing performance. I updated some of the manuscript to reflect these changes.

---
title: "Statistical Analysis Script"
author: "Taylor Glass and Arlyn Santiago"
date: "03/12/24"
output: html_document
---
## Load necessary packages
```{r}
#| message: false
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) # for model building 
```
## Load the cleaned data
```{r}
#path to data
data_location <- here("data","processed-data","processeddata2.rds")
#load data
mydata <- readRDS(data_location)
```

## Data fitting and statistical analysis 
The first trend we will investigate is if there is a stronger relationship between total usage of one method of birth control and percentage of unintended pregnancies associated with that method across the three birth control methods and no usage options. These models will allow us to predict unintended pregnancies based on population data about number of women using each type of contraception method per country.

### First Model Fit 
The first model attempts to predict the percentage of unintended pregnancies among women using long acting reversible  and sterilization methods of contraception, 'pct_upreg_larcster', based on the total number of women using this method of contraception. We will use the version of the variable that has been standardized by population size, larcster_standard.
```{r}
# fit linear model using pct_upreg_larcster as outcome, larcster_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit1 <- lm_mod %>% 
            fit(pct_upreg_larcster ~ larcster_standard, mydata) #estimate/train the model

@@

# save fit results table  
table_file1 = here("results", "resulttable1.rds")
saveRDS(lmtable1, file = table_file1)
```

### Second Model Fit 
The second model attempts to predict the percentage of unintended pregnancies among women using short acting reversible methods of contraception 'pct_upreg_sarc' based on the total number of women using this method of contraception. We will use the version of the variable that has been standardized by population size, 'sarc_standard'. 
```{r}
# fit linear model using pct_upreg_sarc as outcome, sarc_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit2 <- lm_mod %>% 
            fit(pct_upreg_sarc ~ sarc_standard, mydata) #estimate/train the model

@@

# save fit results table  
table_file2 = here("results", "resulttable2.rds")
saveRDS(lmtable2, file = table_file2)
```
### Third Model Fit 
The second model attempts to predict the percentage of unintended pregnancies among women using traditional of contraception 'pct_upreg_trad' based on the total number of women using this method of contraception. We will use the version of the variable that has been standardized by population size, 'trad_standard'. 
```{r}
# fit linear model using pct_upreg_trad as outcome, trad_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit3 <- lm_mod %>% 
            fit(pct_upreg_trad ~ trad_standard, mydata) #estimate/train the model

# generate clean output with estimates and p-values
tidy(lmfit3)

@@

# save fit results table  
table_file3 = here("results", "resulttable3.rds")
saveRDS(lmtable3, file = table_file3)
```
### Fourth Model Fit 
The fourth model attempts to predict the percentage of unintended pregnancies among women using no method of contraception 'pct_upreg_nouse' based on the total number of women not using a method of contraception. We will use the version of the variable that has been standardized by population size, 'none_standard'. 
```{r}
# fit linear model using pct_upreg_nouse as outcome, none_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit4 <- lm_mod %>% 
            fit(pct_upreg_nouse ~ none_standard, mydata) #estimate/train the model

# generate clean output with estimates and p-values
tidy(lmfit4)

@@

# save fit results table  
table_file4 = here("results", "resulttable4.rds")
saveRDS(lmtable4, file = table_file4)
```
Based on the coefficient estimates for each parameter, the association between total number of women not using a contraception method is most strongly correlated with percentage of unintended pregnancies among those not using contraceptive methods, followed by those who use traditional methods and short-acting reversible methods. The long acting reversible and sterilization methods shows the weakest association based on these four models.

We want to explore if a model with all the variables of total number of women using each type of contraceptive method would be more accurate in estimating percentage of unintended pregnancies.

### Fifth Model Fit 
```{r}
# fit linear model using pct_upreg_nouse as outcome, none_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit5 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_larcster ~ larcster_standard + sarc_standard + trad_standard + none_standard, mydata)

@@

# save fit results table  
table_file5 = here("results", "resulttable5.rds")
saveRDS(lmtable5, file = table_file5)
```
These results show that only 'larcster_standard' showed a positive relationship with percentage of unintended pregnancies associated with long acting reversible and standardization techniques. This is to be expected.

We also want to determine if there is a relationship between the region and percentage of unintended pregnancies associated with each method of contraception.

### Sixth Model Fit 
```{r}
# fit linear model using pct_upreg_nouse as outcome, none_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit6 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_larcster ~ region, mydata)

@@

# save fit results table  
table_file6 = here("results", "resulttable6.rds")
saveRDS(lmtable6, file = table_file6)
```

>>>>

---
title: "Statistical Analysis Script"
author: "Taylor Glass and Arlyn Santiago"
date: "03/12/24"
output: html_document
---
## Load necessary packages
```{r}
#| message: false
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) # for model building 
```

## Load the cleaned data
```{r}
#path to data
data_location <- here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

## Data fitting and statistical analysis 
The outcome of interest in all of the following models is the percentage of unintended pregnancies associated with usage of short acting reversible contraception methods. The predictors used in our models include total number of women using short acting reversible methods standardized by population size, region, percent of women currently married, percent of women never married, total rate of maternal deaths, rate of maternal deaths due to unintended pregnancy, and rate of maternal deaths due to abortion. We are considering adding another predictor using a political freedom index for each country.

### First Model Fit 
The second model attempts to predict the percentage of unintended pregnancies among women using short acting reversible methods of contraception 'pct_upreg_sarc' based on the total number of women using this method of contraception. We will use the version of the variable that has been standardized by population size, 'sarc_standard'. 
```{r}
# fit linear model using pct_upreg_sarc as outcome, sarc_standard as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit1 <- lm_mod %>% 
            fit(pct_upreg_sarc ~ sarc_standard, mydata) #estimate/train the model

@@

# save fit results table  
table_file1 = here("results", "resulttable1.rds")
saveRDS(lmtable1, file = table_file1)
```
There appears to be a very strong relationship between total number of women using this birth control method and the outcome based on the coefficient estimate of 87.65998.

### Second Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, region as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit2 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_larcster ~ region, mydata)

@@

# save fit results table  
table_file2 = here("results", "resulttable2.rds")
saveRDS(lmtable2, file = table_file2)
```
This model shows that the Latin America/the Carribean region has the strongest association with the outcome variable based on a coefficient estimate of 1.6218844 followed closely by the Asia region. The Oceania region has the weakest association with the outcome variable based on a coefficient estimate of 0.2512100.

### Third Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, pct_currentlymarried as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit3 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ pct_currentlymarried, mydata)

#generate clean output with estimates and p-values
tidy(lmfit3)

@@

# save fit results table  
table_file3 = here("results", "resulttable3.rds")
saveRDS(lmtable3, file = table_file3)
```
This model shows there is a negative relationship between the number of women currently married and the outcome. As the percentage of currently married women increases by 1, the percentage of unintended pregnancies decreases by a factor of 0.4181571.

# Fourth Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, pct_nevermarried as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit4 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ pct_nevermarried, mydata)

#generate clean output with estimates and p-values
tidy(lmfit4)

@@

# save fit results table  
table_file4 = here("results", "resulttable4.rds")
saveRDS(lmtable4, file = table_file4)
```
This model shows a positive relationship between the percentage of women never married and the outcome. As the percentage of women never married increases by 1, the percentage of unintended pregnancies increases by a factor of 0.2474218.

# Fifth Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, rate_matdeaths as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit5 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ rate_matdeaths, mydata)

@@

# save fit results table  
table_file5 = here("results", "resulttable5.rds")
saveRDS(lmtable5, file = table_file5)
```
This model shows that total rate of maternal deaths is slightly negatively correlated with the outcome based on a small coefficient estimate of -0.01482568.

# Sixth Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, rate_matdeaths_upreg as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit6 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ rate_matdeaths_upreg, mydata)

@@

# save fit results table  
table_file6 = here("results", "resulttable6.rds")
saveRDS(lmtable6, file = table_file6)
```
This model shows that total rate of maternal deaths due to unintended pregnancies has an even weaker association with the outcome based on a small coefficient estimate of -0.003158542.

# Seventh Model Fit 
```{r}
# fit linear model using pct_upreg_sarc as outcome, rate_matdeaths_abs as predictor
lm_mod <- linear_reg() #specify the type of model
lmfit7 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ rate_matdeaths_abs, mydata)

#generate clean output with estimates and p-values
tidy(lmfit7)

# save results from fit into a data frame 
lmtable7 <- tidy(lmfit7)

# save fit results table  
table_file7 = here("results", "resulttable7.rds")
saveRDS(lmtable7, file = table_file7)
```
This model shows that total rate of maternal deaths due to abortion has the strongest association with the outcome out of the three maternal mortality rates included based on a coefficient estimate of -0.033353527.

### Eighth Model Fit 
We will explore all 7 predictors using in a multiple linear regression model to determine how the associations  interact. 
```{r}
# fit linear model using pct_upreg_sarc as outcome with all 7 predictors
lm_mod <- linear_reg() #specify the type of model
lmfit8 <- lm_mod %>% #estimate/train the model 
            fit(pct_upreg_sarc ~ sarc_standard + region + pct_currentlymarried + pct_nevermarried + rate_matdeaths + rate_matdeaths_upreg + rate_matdeaths_abs, mydata)

#generate clean output with estimates and p-values
tidy(lmfit8)

# save results from fit into a data frame 
lmtable8 <- tidy(lmfit8)

# save fit results table  
table_file8 = here("results", "resulttable8.rds")
saveRDS(lmtable8, file = table_file8)
```
This model shows the strongest predictor is the number of women using short acting reversible contraception with a huge coefficient of 85.6433. The next strongest predictor is the European region followed by the Latin America/the Caribbean region and the Asia region, which is a shift from the model with the singular region predictor. The Oceania region now has a negative association with the outcome. Both of the marriage status predictors are associated with a decrease in the outcome. The coefficient estimates for the rate of maternal death due to unintended pregnancies and abortion both flipped from the single predictor models to show a positive association with the outcome.

## Assessing Model Performance
It is important to consider appropriate metrics to measure model performance. We will use RMSE at this point to compare the model using the number of women using short acting reversible contraception predictor and the model using the percent currently married predictor. 
```{r}
# create predictions using the sarc_standard predictor model
poppred <- predict(lmfit1, new_data = mydata) %>% 
                  select(.pred)

# create a data frame with the predictions and true values of Y
popdata <- bind_cols(poppred, mydata$sarc_standard) %>%
            rename(sarc_standard = "...2")

# find RMSE and R squared to determine model fit
rmse1 <- rmse(popdata, truth = sarc_standard, estimate = .pred)

# create predictions using the pct_currentlymarriedd predictor model
marriedpred <- predict(lmfit3, new_data = mydata) %>% 
                  select(.pred)

# create a data frame with the predictions and true values of Y
marrieddata <- bind_cols(marriedpred, mydata$pct_currentlymarried) %>%
            rename(pct_currentlymarried = "...2")

# find RMSE and R squared to determine model fit
rmse2 <- rmse(marrieddata, truth = pct_currentlymarried, estimate = .pred)

# compare the metrics 
print(rmse1)
print(rmse2)
```

The RMSE metric favors the model with the total number of women using the short acting reversible contraception method over the model with the percentage of women currently married, which is to be expected.

### 2024-03-14 products/manuscript/Manuscript.qmd
Statistical Analysis Update
I cleaned this section to only include the one outcome we chose with models for each of the 7 predictors. I also added RMSE values for 2 models to begin comparing performance. I updated some of the manuscript to reflect these changes.

# Summarize the data 
```{r}
table(AIUdata$region)
table(AIUdata$subregion)
summary(AIUdata$pct_upreg)
ggplot(AIUdata, aes(x = country, y = pct_upreg, fill = region)) +
  geom_col(position = "dodge") +
  labs(title = "Percentage Unintended Pregnancies by Country",
       x = "Country",
       y = "Percentage")
```
The dataset includes 53 African countries, 36 Asian countries, 11 European Countries, 24 Latin American/Caribbean countries, and 8 Oceanic countries.There are 16 subregions that provide better detail about each country's location. The average percentage of unintended pregnancies at the country level is 46.79%. with a maximum of 75.65%. This variable has 24 missing values, which will need to be cleaned.  A simple visualization shows that African and Latin American/Carribean countries have the highest percentages of unintended pregnancies.

# Research Questions
1) Is there a significant difference in the percentage of unplanned pregnancies between contraceptive methods among all countries? Does this difference vary by region?
The main outcome variables will be total percentage of unintended pregnancies per birth control method (`pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`). 
The control variable will be total percentage of unintended pregnancies among women not using birth control (`pct_upreg_none`)
The predictor variables will include the total number of people women using each type of birth control standardized by population size, which will be utilized to create a factor variable showing low, medium, and high usage for each method per country. 
Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. These variables will be combined across age groups to create a total usage variable for each method. 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

We will be looking for patterns in contraceptive methods and higher unintended pregnancies across age groups to investigate the effectiveness of each method. We hope to find relationships between subregions, rates of unintended pregnancies, and contraceptive methods to determine possible interventions that will increase the effectiveness of contraceptive methods.

# How to analyze 
All of the variables used in this analysis are numeric, so we can build linear regression models to analyze how each method of contraception affects the percentage of unintended pregnancies. We can group countries together based on the subregion variable to compare how these relationships vary across the world as shown in the crude visualization above. There are other interesting variables that can be used for grouping for analysis such as wealth, urban/rural location, and marriage status.

>>>>

# Summarize the data 
```{r}
table(AIUdata$region)
table(AIUdata$subregion)
summary(AIUdata$pct_upreg)
ggplot(AIUdata, aes(x = country, y = pct_upreg, fill = region)) +
  geom_col(position = "dodge") +
  labs(title = "Percentage Unintended Pregnancies by Country",
       x = "Country",
       y = "Percentage")
```
The dataset includes 53 African countries, 36 Asian countries, 11 European Countries, 24 Latin American/Caribbean countries, and 8 Oceanic countries.There are 16 subregions that provide better detail about each country's location. The average percentage of unintended pregnancies at the country level is 46.79%. with a maximum of 75.65%. This variable has 24 missing values, which will need to be cleaned.  A simple visualization shows that African and Latin American/Caribbean countries have the highest percentages of unintended pregnancies.

# Research Questions
1) Is there a significant difference in the percentage of unintended pregnancies among short acting reversible contraception users among countries? 
    The main outcome variables will be total percentage of unintended pregnancies among women using short acting          reversible contraception methods (`pct_upreg_sarc`).

    The predictor variables will include the total number of people women using each type of birth control                standardized by population size, region, percentage of women currently married, percentage of women never married,     total rate of maternal deaths, rate of maternal deaths due to unintended pregnancy and rate of maternal deaths due     to abortion.

    A full list of variables can be found in the processing-file in the code > processing-code folder.

# How to analyze 
Most of the predictors used in this analysis are numeric, so we can build linear regression models to analyze how each predictor affects the percentage of unintended pregnancies among women using short acting reversible methods. We can group countries together based on the subregion variable to compare how these relationships vary across the world as shown in the crude visualization above. There are other interesting variables that can be used for grouping for analysis such as wealth, urban/rural location, and marriage status.
